---
title: "Advanced Dynamics and Temporal Cascades"
author: "Masahiro Ono"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Dynamics and Temporal Cascades}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(CanonicalTockySeq)

```

## Introduction

Once the **Canonical Tocky Manifold** is constructed and cells are assigned a precise continuous coordinate (Tocky Time), the next crucial step is determining which genes significantly alter their expression along this developmental trajectory.

This vignette demonstrates how to use `RunTradeSeqTocky` to fit Generalized Additive Models (GAMs) and `PlotTockyHeatmap` to identify transcriptional cascades.

## 1. Establishing the Base Manifold
We will simulate a dataset, ensuring the expression matrix represents non-negative integer counts, and assign the cells to two biological conditions (WT and KO).

```{r simulate-and-build}
set.seed(123)
n_genes <- 500  # Increased from 100 to generate a richer dataset
n_cells <- 300

# 1. Simulate Constraints
Z <- matrix(runif(n_genes * 3), nrow = n_genes, ncol = 3)
colnames(Z) <- c("Blue", "BlueRed", "Red")
rownames(Z) <- paste0("Gene_", 1:n_genes)

# 2. Simulate Expression as Non-Negative Counts
cell_loadings <- matrix(runif(3 * n_cells), nrow = 3, ncol = n_cells)
X <- round(Z %*% cell_loadings * 10 + matrix(rnorm(n_genes * n_cells, sd = 2), nrow = n_genes))
X[X < 0] <- 0 
colnames(X) <- paste0("Cell_", 1:n_cells)
rownames(X) <- rownames(Z)

# 3. Assign Two Conditions (WT and KO)
cell_groups <- sample(c("WT", "KO"), n_cells, replace = TRUE)
names(cell_groups) <- colnames(X)

# 4. Reconstruct Manifold
tocky_res <- CanonicalTockySeq(X = X, Z = Z)
gradient_res <- GradientTockySeq(
  res = tocky_res, 
  B = tocky_res$biplot["Blue",], 
  BR = tocky_res$biplot["BlueRed",], 
  R = tocky_res$biplot["Red",]
)
```

## 2. Rapid Pre-filtering of Dynamic Genes

Fitting GAMs to tens of thousands of genes is computationally intensive. The `SelectTockyGenes` function provides a rapid linear spline regression filter to isolate the most dynamically varying genes along the Tocky Time axis.

```{r pre-filter}
# Select the top 100 most dynamic genes for a highly impressive cascade
dynamic_genes <- SelectTockyGenes(
  expression_data = X, 
  gradient_res = gradient_res, 
  top_n = 100, 
  min_expr = 0.05
)

```

## 3. Differential Dynamics using tradeSeq
The `RunTradeSeqTocky` wrapper interfaces directly with the `tradeSeq` package to fit GAMs. It evaluates expression against the 0-90 degree Tocky Time. 

By providing a named vector with two conditions (WT and KO), `tradeSeq` will test for genes that have significantly different dynamic trajectories between the two groups.

```{r run-tradeseq}
# Fit GAMs and perform statistical testing between WT and KO
tradeSeq_res <- RunTradeSeqTocky(
  object = X,
  gradient_res = gradient_res,
  group_by = cell_groups,
  genes = dynamic_genes,
  n_knots = 5,
  n_cores = 1
)

# View the differential dynamics results (p-values and FDR)
head(tradeSeq_res$results)
```

## 4. Visualizing the Temporal Cascade

Once significant genes are identified, plotting them in developmental order reveals the sequential transcriptional logic of your system.

The `PlotTockyHeatmap` function supports ordering genes by their peak activation timing ("peak") or via standard hierarchical clustering ("cluster"). By default, it applies a Z-score row scaling to highlight expression shifts.

```{r plot-cascade, fig.width=7, fig.height=5}
# Plot the expression cascade, ordered by the peak Tocky Time
PlotTockyHeatmap(
  object = X,
  gradient_res = gradient_res,
  genes = dynamic_genes,
  ordering_method = "peak",
  n_bins = 50,
  span = 0.5
)

```

## 5. Biological Interpretation

The heatmap above visualizes the transcriptional cascade along the Canonical Tocky Manifold. 

* **The X-axis (Tocky Time):** Represents the progression of cells from the 'New' state to the 'Arrested' state, decoupled from the absolute signaling intensity.
* **The Y-axis (Genes):** Displays the top 100 most dynamic genes, ordered sequentially by their peak expression timing.
* **The Diagonal Pattern:** A smooth, continuous diagonal band indicates a highly coordinated developmental program where genes are activated and silenced in a strict temporal order.

By integrating `CanonicalTockySeq` with `tradeSeq`, you can isolate these specific temporal windows, identifying the immediate-early genes that initiate differentiation versus the late-stage markers associated with terminal states.