---
title: "Getting Started with CanonicalTockySeq"
author: "Masahiro Ono"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CanonicalTockySeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(CanonicalTockySeq)

```

## Introduction

**CanonicalTockySeq** utilizes a supervised Canonical Redundancy Analysis (RDA) to construct a transcriptomic manifold with conical geometry. This vignette demonstrates the core workflow: generating the canonical space, projecting cells via piecewise SLERP, and analyzing temporal gene dynamics.

## 1. Simulating Toy Data

To demonstrate the workflow, we will first generate a synthetic single-cell dataset.

The `CanonicalTockySeq` function requires two inputs:

* 
**`X`**: A numeric matrix of gene expression (Genes x Cells).


* 
**`Z`**: A numeric matrix of explanatory variables (Genes x Signatures), typically representing Tocky-specific gene signatures like Blue, Blue-Red, and Red.



```{r simulate-data}
set.seed(2026)

# Define dimensions
n_genes <- 100
n_cells <- 500

# 1. Simulate the Constraint Matrix Z (100 Genes x 3 Signatures)
Z <- matrix(runif(n_genes * 3), nrow = n_genes, ncol = 3)
colnames(Z) <- c("Blue", "BlueRed", "Red")
rownames(Z) <- paste0("Gene_", 1:n_genes)

# 2. Simulate the Expression Matrix X (100 Genes x 500 Cells)
# We multiply Z by random cell loadings to embed the biological trajectory, 
# then add some statistical noise.
cell_loadings <- matrix(runif(3 * n_cells), nrow = 3, ncol = n_cells)
X <- Z %*% cell_loadings + matrix(rnorm(n_genes * n_cells, sd = 0.5), nrow = n_genes)
colnames(X) <- paste0("Cell_", 1:n_cells)
rownames(X) <- rownames(Z)

```

## 2. Canonical Redundancy Analysis (RDA)

Now we apply the RDA architecture to our simulated data, treating genes as "sites" and cells as "species" constrained by the Tocky stages.

```{r run-rda}
# Perform the RDA procedure
tocky_res <- CanonicalTockySeq(X = X, Z = Z)

# The result contains cell_scores and biplot vectors representing the constraints
head(tocky_res$cell_scores)

```

## 3. Manifold Reconstruction via SLERP

To resolve the temporal gradient, we project the cells onto a piecewise Spherical Linear Interpolation (SLERP) manifold. We define the landmarks based on the biplot vectors from our RDA.

```{r run-slerp}
# Extract landmark vectors from the RDA biplot
landmark_B  <- tocky_res$biplot["Blue", ]
landmark_BR <- tocky_res$biplot["BlueRed", ]
landmark_R  <- tocky_res$biplot["Red", ]

# Calculate Tocky Time (Angle) and Tocky Intensity (Radial norm)
gradient_res <- GradientTockySeq(
  res = tocky_res, 
  B = landmark_B, 
  BR = landmark_BR, 
  R = landmark_R, 
  filter_negative = TRUE
)

head(gradient_res)

```

## 4. Visualizing the Conical Manifold

We can now visualize the constrained ordination space. The `plotCanonicalTocky` function displays the axes alongside a continuous gradient legend for Tocky Time.

```{r plot-manifold, fig.width=8, fig.height=4}
plotCanonicalTocky(tocky_res, gradient_res, alpha_level = 0.5)

```

## 5. Identifying Gene Dynamics

With the continuous temporal gradient established, we can rapidly filter for genes that vary significantly along the Tocky Time axis.

```{r gene-dynamics}
# Filter for top dynamic genes
dynamic_genes <- SelectTockyGenes(
  expression_data = X, 
  gradient_res = gradient_res, 
  top_n = 5
)

# Plot the expression dynamics of the top gene using LOESS smoothing
plotGeneDynamics(
  expression_data = X, 
  gradient_res = gradient_res, 
  gene = dynamic_genes[1], 
  span = 0.8
)

```
