---
title: "Getting Started with CanonicalTockySeq"
author: "Masahiro Ono"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CanonicalTockySeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(CanonicalTockySeq)

```

## Introduction

**CanonicalTockySeq** utilizes a supervised Canonical Redundancy Analysis (RDA) to construct a transcriptomic manifold with conical geometry. By anchoring single-cell expression data to known developmental landmarks, it allows researchers to accurately trace continuous *in vivo* trajectories.

This vignette demonstrates the core workflow: simulating a structured transcriptomic manifold, generating the canonical space, projecting cells via piecewise SLERP, and visualizing the resulting trajectory.

## 1. Simulating a Transcriptomic Manifold

To demonstrate the workflow, we will generate a synthetic single-cell dataset with a true underlying biological structure. We engineer continuous temporal cascades using Gaussian functions so that genes peak sequentially along a developmental trajectory, mirroring real biological differentiation.

The `CanonicalTockySeq` function requires two inputs:

* **`X`**: A numeric matrix of gene expression (Genes x Cells).
* **`Z`**: A numeric matrix of explanatory variables (Genes x Signatures), representing Tocky-specific landmark signatures.

```{r simulate-data}
set.seed(2026)

n_genes <- 500
n_cells <- 300

# 1. Assign true developmental time to cells
true_time <- runif(n_cells, 0, pi/2)
cell_names <- paste0("Cell_", 1:n_cells)

# 2. Engineer Continuous Temporal Cascades for Top 100 Genes
# We use a Gaussian function so each gene peaks sequentially along the trajectory
gene_peaks <- seq(0, pi/2, length.out = 100)
X_signal <- matrix(0, nrow = 100, ncol = n_cells)
for(i in 1:100) {
  X_signal[i, ] <- exp( - (true_time - gene_peaks[i])^2 / 0.1) * 20
}

# Add 400 random noise genes to mimic background transcription
X_noise <- matrix(runif(400 * n_cells, 0, 2), nrow = 400, ncol = n_cells)
X_base <- rbind(X_signal, X_noise)

# Add statistical noise and format as non-negative counts
X <- round(X_base + matrix(rnorm(n_genes * n_cells, sd = 1), nrow = n_genes))
X[X < 0] <- 0
colnames(X) <- cell_names
rownames(X) <- paste0("Gene_", 1:n_genes)

# 3. Create Gene Constraints (Z) matching the biological landmarks
# Blue (New) peaks at t=0, BlueRed (Persistent) at t=pi/4, Red (Arrested) at t=pi/2
all_gene_peaks <- c(gene_peaks, runif(400, 0, pi/2))
Z <- matrix(0, nrow = n_genes, ncol = 3)
colnames(Z) <- c("Blue", "BlueRed", "Red")
rownames(Z) <- rownames(X)

Z[, "Blue"]    <- exp( - (all_gene_peaks - 0)^2 / 0.2 )
Z[, "BlueRed"] <- exp( - (all_gene_peaks - pi/4)^2 / 0.2 )
Z[, "Red"]     <- exp( - (all_gene_peaks - pi/2)^2 / 0.2 )

```

## 2. Canonical Redundancy Analysis (RDA)

Now we apply the RDA architecture to our structured data, treating genes as "sites" and cells as "species" constrained by the Tocky stages.

```{r run-rda}
# Perform the RDA procedure
tocky_res <- CanonicalTockySeq(X = X, Z = Z)

# The result contains cell_scores and biplot vectors representing the constraints
head(tocky_res$cell_scores)

```

## 3. Manifold Reconstruction via SLERP

To resolve the temporal gradient, we project the cells onto a piecewise Spherical Linear Interpolation (SLERP) manifold. We define the landmarks based on the biplot vectors extracted directly from our RDA.

```{r run-slerp}
# Extract landmark vectors from the RDA biplot
landmark_B  <- tocky_res$biplot["Blue", ]
landmark_BR <- tocky_res$biplot["BlueRed", ]
landmark_R  <- tocky_res$biplot["Red", ]

# Calculate Tocky Time (Angle) and Tocky Intensity (Radial norm)
gradient_res <- GradientTockySeq(
  res = tocky_res, 
  B = landmark_B, 
  BR = landmark_BR, 
  R = landmark_R, 
  filter_negative = TRUE
)

head(gradient_res)

```

## 4. Visualizing the Conical Manifold

We can now visualize the constrained ordination space. Notice how the simulated temporal cascades naturally form a sweeping, continuous biological trajectory. The `plotCanonicalTocky` function displays the axes alongside a continuous gradient legend for Tocky Time.

```{r plot-manifold, fig.width=9, fig.height=4.5}
plotCanonicalTocky(tocky_res, gradient_res, alpha_level = 0.5)

```

## 5. Identifying Gene Dynamics

With the continuous temporal gradient established along our arched manifold, we can rapidly filter for genes that vary significantly along the Tocky Time axis.

```{r gene-dynamics}
# Filter for top dynamic genes
dynamic_genes <- SelectTockyGenes(
  expression_data = X, 
  gradient_res = gradient_res, 
  top_n = 5
)

# Plot the expression dynamics of a top gene using LOESS smoothing
plotGeneDynamics(
  expression_data = X, 
  gradient_res = gradient_res, 
  gene = dynamic_genes[1], 
  span = 0.8
)

```
---
